---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


```{r, include = FALSE}
invisible(lapply(list.files("R/", full.names = TRUE), source))
library(MplusAutomation)
```

# DSEM2R

<!-- badges: start -->
<!-- badges: end -->

# DSEM2R

This repository includes a set of functions that allow to easily run DSEM analyses within the R environment. In particular, with these functions one can run (V)AR and multilevel (V)AR models. The functions allow writing the syntax and exporting the data to Mplus, run the model in Mplus, and read the output of the analysis.

**Note that you need to have Mplus installed to make use of the functions available in this repository as well as the R package [MplusAutomation](https://cran.r-project.org/web/packages/MplusAutomation/index.html).**

There are two main functions: `var2Mplus` and `mlvar2Mplus`. `var2Mplus` is used to run $N=1$ DSEM models and `mlvar2Mplus` is used to run multilevel DSEM models.

## How to Use It

At the moment, the best way to make use of these functions is to download all the files in the folder [R](R/) and `source` the functions to your R environment. In the future, we plan to structure this repository as an R package so users can simply install the package via `remotes::install_github("secastroal/DSEM2R")`.

### Example: var2Mplus

Here, we present some examples on how to use the function `var2DSEM`.

Let us first simulate time series data with two variables and 100 time points based on a VAR(1) process:

```{r}
vardata <- freqdom::rar(100)
vardata <- data.frame(vardata)
names(vardata) <- c("y1", "y2")
```

#### AR model.

Now, with the simulated data, we can simply fit an AR(1) model for variable `y1`. To do this, three arguments are needed: `y` to indicate the name of the dependent variable (`y1`), `data` to indicate the data.frame where the data is stored, and `filename` to indicate the name for the data and input files that are required to run the analysis in Mplus. When running the function, it prints the Mplus syntax of the model in the console and at the same time it has run the model and read the results into R. By default, the function also saves the MCMC samples into a *.dat* file and uses as default output the options *TECH1* and *TECH8* from Mplus.

```{r, message=FALSE, warning=FALSE}
exvar01 <- var2Mplus(
  y        = "y1",
  data     = vardata,
  filename = "exvar01.dat")
```

 
The final object `exvar01` is an *"mplus.model"* object as defined in `MplusAutomation`, which includes all the available output of the model. For example, we can use the following code to access the parameter estimates:

```{r}
exvar01$parameters$unstandardized
```

Also, we can extract the MCMC samples. the following command extracts the first 6 samples of the first chain:

```{r}
head(exvar01$bparameters$valid_draw$'1')
```

#### AR model with one covariate.

Now, we can also fit an AR model on variable `y1` but adding `y2` as a covariate. By default, when adding covariates, the function includes the contemporaneous and the lagged effect of the covariate on the dependent variable. This behavior can be modified with the argument `beta.at.0`. More information about this will be available in the documentation. For now, if the model is defined as:

$$
y_{1,t} = \alpha + \phi{y_{1,t-1}} + \beta_{0}{y_{2,t}} + \beta_{1}{y_{2,t - 1}} + \varepsilon_{t},
$$
then, the syntax to run the model is Mplus is as follows:

```{r, message=FALSE, warning=FALSE}
exvar02 <- var2Mplus(
  y        = "y1",
  x        = "y2",
  data     = vardata,
  filename = "exvar02.dat")
```

#### VAR model.

In this example, we use `y2` as a dependent variable instead of as a covariate.Therefore, we fit a bivariate VAR model. We can specify this model with the following command: 

```{r, message=FALSE, warning=FALSE}
exvar03 <- var2Mplus(
  y        = c("y1", "y2"),
  data     = vardata,
  filename = "exvar03.dat")
```

#### Additional options.

With the function `var2Mplus`, we can further customize the analyses in Mplus. For example, we can specify a time variable to use with the *TINTERVAL* option, modify the settings of the MCMC algorithm, or save the factor scores of the latent variables of interest. The example bellow shows how to modify the setting of the MCMC algorithm:

```{r, message=FALSE, warning=FALSE}
exvar04 <- var2Mplus(
  y        = c("y1", "y2"),
  data     = vardata,
  filename = "exvar04.dat",
  analysis_options = list(chains = 4,
                          biterations.min = 2000,
                          biterations.max = 10000,
                          thin = 10))

```

### Example: mlvar2Mplus

In contrast, the function `mlvar2Mplus` is used to fit multilevel VAR models. 

Again, let us first simulate a data set with 50 participants, 3 variables, and 100 time points:

```{r}
mlvardata <- mlVAR::mlVARsim(nPerson = 50, 
                             nNode   = 3, 
                             nTime   = 100)
mlvardata <- mlvardata$Data
names(mlvardata) <- c("y1", "y2", "y3", "id")
```

#### ML-AR model

Using `mlvar2Mplus` is very similar to using `var2Mplus` but there are much more options to accommodate for the multilevel structure (occasions nested in individuals). This function allows adding within-level covariates, between-level covariates, and between-level dependent variables. Also, by default, lagged coefficients, regression slopes, linear time trends, and residual variances are allowed to vary randomly across individuals.

In this first example, we simply fit a multilevel AR model for one variable:

```{r, message=FALSE, warning=FALSE}
exmlvar01 <- mlvar2Mplus(
  y    = "y1",
  id   = "id",
  data = mlvardata,
  filename = "exmlvar01.dat")
```

Just as with `var2Mplus`, the object `exmlvar01` is an *"mplus.model"* object. Hence, we can look at the estimated parameters like this:

```{r}
exmlvar01$parameters$unstandardized
```

And, we can look at the MCMC samples like this:

```{r}
head(exmlvar01$bparameters$valid_draw$'1')
```

#### ML-VAR(1)

We can fit a multilevel VAR to the three simulated variables. However, in this case, we won't alow the residual variances to vary randomly across participants:

```{r, message=FALSE, warning=FALSE}
exmlvar02 <- mlvar2Mplus(
  y    = c("y1", "y2", "y3"),
  id   = "id",
  data = mlvardata,
  random.effects = list(lagged = TRUE, rvar = FALSE),
  filename = "exmlvar02.dat")
```

#### ML-VAR(2)

The order of the lagged effects can also be modified:

```{r, message=FALSE, warning=FALSE}
exmlvar03 <- mlvar2Mplus(
  y    = c("y1", "y2"),
  id   = "id",
  data = mlvardata,
  lags = 2,
  random.effects = list(lagged = TRUE, rvar = FALSE),
  filename = "exmlvar03.dat")
```

#### Additional options.

In `mlvar2Mplus`, we can also further change the settings of the analysis in Mplus, just as with `var2Mplus`.

```{r, message=FALSE, warning=FALSE}
exmlvar04 <- mlvar2Mplus(
  y    = c("y1", "y2"),
  id   = "id",
  data = mlvardata,
  lags = 2,
  random.effects = list(lagged = TRUE, rvar = FALSE),
  filename = "exmlvar04.dat",
  analysis_options = list(chains = 4,
                          biterations.min = 5000,
                          biterations.max = 20000,
                          thin = 10))
```

## To Do
- [ ] Write package Description and documentation.
- [ ] Test installing the package via `devtools`.
- [ ] Allow for contemporaneous effects among dependent variables $y$ in both `var2Mplus` and `mlvar2Mplus`.
- [ ] Include syntax line to allow for within level covariates when TINTERVAL is defined. Trick Mplus into thinking the covariate is a dependent variable. For example: `x ON x&1\@0`.
- [ ] Group center within-level covariates in mlvar2Mplus.
- [ ] Test whether mlvar2DSEM works when only specifying some of the random.effects.
- [ ] Include time in the between var-cov structure when using time for a time trend.
- [ ] Maybe extend to allow for latent variables.
- [ ] Facilitate the exploration of output.

<!-- You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this. You could also use GitHub Actions to re-render `README.Rmd` every time you push. An example workflow can be found here: <https://github.com/r-lib/actions/tree/v1/examples>. -->

