---
title: "ML-VAR Analyses"
author:
  - Sebastian Castro-Alvarez
  - Laura F. Bringmann
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
  bookdown::word_document2:
    fig_caption: yes
    number_sections: FALSE
    reference_docx: APAtemplate.docx
bibliography: references.bib
csl: apa7.csl
link-citations: true
always_allow_html: true
---
  
```{r setup, include=FALSE}
library(knitr)
library(kableExtra)
library(bookdown)
library(pander)
knitr::opts_chunk$set(echo = TRUE, fig.height = 4, fig.width = 6, fig.pos = "!H",  
                      warning=FALSE, message=FALSE)

panderOptions('table.alignment.default', function(df)
  ifelse(sapply(df, is.numeric), 'right', 'left'))
panderOptions('table.split.table', Inf)
panderOptions('big.mark', ",")
panderOptions('keep.trailing.zeros', TRUE)

# Use this code to render the document:
# Render the document to a .docx file.
# rmarkdown::render("Rmarkdown/ML-VAR_Analyses.Rmd") 
```

In this document, we analyze intensive longitudinal data on the emotions sad and interest.

```{r}
if (!"psych" %in% rownames(installed.packages())) {install.packages("psych")}
if (!"naniar" %in% rownames(installed.packages())) {install.packages("naniar")}
if (!"lubridate" %in% rownames(installed.packages())) {install.packages("lubridate")}
if (!"esmpack" %in% rownames(installed.packages())) {
  remotes::install_github("secastroal/esmpack")
  }
library(esmpack)
library(foreign)
```

## Reading Data

```{r}
mlvardata <- read.csv2("deid_cleaned_lifepak_ttt_phase_1.csv", 
                       header = TRUE, sep = ",")
```

The data has 14 variables: ID, response number, beep time, response time, time difference (unclear in relation to what), beep completed, and scores on 6 emotions.

```{r}
summary(mlvardata[, 7:14])
```

The emotions were measured based on a VAS from 0 to 100. However, there are `r sum(mlvardata[, "interest"] < 0, na.rm = TRUE)` observations on the variable interest with scores lower than 0.

The data includes responses from `r nsub(mlvardata$lifepak_id)` participants who replied between `r min(calc.nomiss(interest, lifepak_id, mlvardata))` and `r max(calc.nomiss(interest, lifepak_id, mlvardata))` times. In particular, the distribution of the compliance is presented in the following Figure:

```{r compliance, fig.cap = "Compliance on variable 'interest'"}
hist(calc.nomiss(interest, lifepak_id, mlvardata),
     ylim = c(0, 8),
     xlim = c(0, 120),
     xlab = "Number of complied beeps",
     las = 1,
     breaks = 40,
     main = "")

```

```{r}
par(mfrow = c(1, 2))
hist(mlvardata$interest, 
     main = "",
     las = 1,
     ylim = c(0, 3000),
     xlab = "Interest",
     breaks = 40)
hist(mlvardata$sad,
     main = "",
     las = 1,
     ylim = c(0, 3000),
     xlab = "Sad",
     breaks = 40)
```



