---
title: "ML-VAR Analyses"
author:
  - Sebastian Castro-Alvarez
  - Laura F. Bringmann
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
  bookdown::word_document2:
    fig_caption: yes
    number_sections: FALSE
    reference_docx: APAtemplate.docx
bibliography: references.bib
csl: apa7.csl
link-citations: true
always_allow_html: true
---
  
```{r setup, include=FALSE}
library(knitr)
library(kableExtra)
library(bookdown)
library(pander)
knitr::opts_chunk$set(echo = TRUE, fig.height = 4, fig.width = 6, fig.pos = "!H",  
                      warning=FALSE, message=FALSE)

panderOptions('table.alignment.default', function(df)
  ifelse(sapply(df, is.numeric), 'right', 'left'))
panderOptions('table.split.table', Inf)
panderOptions('big.mark', ",")
panderOptions('keep.trailing.zeros', TRUE)

# Use this code to render the document:
# Render the document to a .docx file.
# rmarkdown::render("Rmarkdown/ML-VAR_Analyses.Rmd") 
```

In this document, we analyze intensive longitudinal data on the emotions sad and interest.

```{r}
if (!"psych" %in% rownames(installed.packages())) {install.packages("psych")}
if (!"naniar" %in% rownames(installed.packages())) {install.packages("naniar")}
if (!"lubridate" %in% rownames(installed.packages())) {install.packages("lubridate")}
if (!"esmpack" %in% rownames(installed.packages())) {
  remotes::install_github("secastroal/esmpack")
  }
library(esmpack)
library(foreign)
```

## Reading Data

```{r}
mlvardata <- read.csv2("deid_cleaned_lifepak_ttt_phase_1.csv", 
                       header = TRUE, sep = ",")
```

The data has 14 variables: ID, response number, beep time, response time, time difference (unclear in relation to what), beep completed, and scores on 6 emotions.

```{r}
summary(mlvardata[, 7:14])
```

The emotions were measured based on a VAS from 0 to 100. However, there are `r sum(mlvardata[, "interest"] < 0, na.rm = TRUE)` observations on the variable interest with scores lower than 0.

```{r}
mlvardata[mlvardata[, "interest"] < 0 & !is.na(mlvardata[, "interest"]), c("lifepak_id", "interest", "sad")]
# remove these 2 subjects
unique(mlvardata[mlvardata[, "interest"] < 0 & !is.na(mlvardata[, "interest"]), c("lifepak_id", "interest", "sad")][, 1]
)

negdata <- mlvardata[mlvardata$lifepak_id == 230042 | mlvardata$lifepak_id == 682098, c("lifepak_id", "interest", "sad")]

plot.ts(negdata$interest[negdata$lifepak_id == 230042])
plot.ts(negdata$interest[negdata$lifepak_id == 682098])
```


```{r}
head(mlvardata[mlvardata$completed_session == 1 & is.na(mlvardata$sad),])
head(mlvardata[mlvardata$completed_session == 0,])
```


The data includes responses from `r nsub(mlvardata$lifepak_id)` participants who replied between `r min(calc.nomiss(interest, lifepak_id, mlvardata))` and `r max(calc.nomiss(interest, lifepak_id, mlvardata))` times. In particular, the distribution of the compliance is presented in the following Figure:

```{r compliance, fig.cap = "Compliance on variable 'interest'"}
hist(calc.nomiss(interest, lifepak_id, mlvardata),
     ylim = c(0, 8),
     xlim = c(0, 120),
     xlab = "Number of complied beeps",
     las = 1,
     breaks = 40,
     main = "")

```

```{r}
par(mfrow = c(1, 2))
hist(mlvardata$interest, 
     main = "",
     las = 1,
     ylim = c(0, 3000),
     xlab = "Interest",
     breaks = 40)
hist(mlvardata$sad,
     main = "",
     las = 1,
     ylim = c(0, 3000),
     xlab = "Sad",
     breaks = 40)
```

The following code confirms the histograms are correct.

```{r}
sum(mlvardata$sad == 0, na.rm = TRUE)
sum(mlvardata$sad == 50, na.rm = TRUE)
sum(mlvardata$sad == 100, na.rm = TRUE)

sum(mlvardata$interest == 0, na.rm = TRUE)
sum(mlvardata$interest == 50, na.rm = TRUE)
sum(mlvardata$interest == 100, na.rm = TRUE)
```

```{r}
mlvardata_60plus <- mlvardata[!(mlvardata$lifepak_id %in% c(230042, 692098)), ]

mlvardata_60plus$ind_nomiss <- ifelse(is.na(mlvardata_60plus$interest) & is.na(mlvardata_60plus$sad), NA, 1)

complianceboth <- calc.nomiss(ind_nomiss, lifepak_id, mlvardata_60plus)

names(which(complianceboth >= 60))

mlvardata_60plus <- mlvardata_60plus[mlvardata_60plus$lifepak_id %in% as.numeric(names(which(complianceboth >= 60))), ]

```


Fit lm with trend (check kpss) and if it is significant plot time series of those individuals

check if there is enough variance... what is considered enough variance? sd of minimum 10% of the scale... meaning minimum sd must be 10.

Plot all the time series per person more than 60 time points.

For multilevel include everybody that has more than 3 observations.
